--- 
title: "Data Analysis and Statistics with R"
subtitle: "Ecological statistics. A collection of statistical resources and courses"
author: "Anthony Davidson; Dylan Z. Childs; others"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output:
  bookdown::gitbook:
    includes:
      in_header: extras.css
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: davan690/stats-for-bio
description: "A Course book for Data Analysis and Statistics with R extended from the Department of Animal and Plant Sciences, University of Sheffield. Credit: Dylan Z. Childs"
---


```{r}
# myPaths <- .libPaths("C:/Program Files/R/R-4.0.1/library")
# myPaths <- c(myPaths)
# .libPaths(myPaths)  # add new path
# .libPaths()
```

**Anthony Davidson:** *July 2020*

I have began building out this project to incorperate some tutoring of statistics for high school students. The core components of these assessments are focused around:

**Linear Regression**

**Anthony Davidson:** *December 2019* I have been using this as a resource for my PhD work and other small projects using biological statistics. The orginal resource can be found at David's site [here](). I have not taught this class but use it as a resource and a place to keep my notes In the open source framework these notes may help develop a ecology and Statistics course.

# Resource information and overview

This is a online course book developed from the __Data Analysis and Statistics with R__ ([APS 240](https://www.shef.ac.uk/aps/currentug/level2/aps240)) module with the extended comments and application from my research and personal projects (Anthony Davidson). 

The raw code for each chapter of the book can be found in the `.Rmd` files in the root folder of the project (`./`).

```{r}
list.files(pattern = "Rmd")
```

### View Book

To view the book you can open the following `.index.Rmd` file:

```{r}

```

You can view this book in any modern desktop browser, as well as on your phone or tablet device. The site is self-contained---it contains all the material you are expected to learn this year.
